Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-13 main.Rmd
acknowledgment part shifted to end

The ideas presented in this article have been implemented in the open-source R [@R-language] package gravitas [@R-gravitas], available from CRAN. More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.

@@

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a _bottom granularity_ and calendar algebra operations are designed to generate new granularities recursively from the bottom up. Some relevant calendar algebra operations are discussed below; these will be used in Section \ref{sec:cyclic-gran} for illustrations in circular and aperiodic cyclic granularities. For a coarser granularity $G_2$ and a finer one $G_1$, a _grouping operation_ partitions some granules of $G_1$  to form a granule of $G_2$. For example, $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$. The _altering-tick operation_ generates new full-integer labeled granularity by periodically expanding or shrinking granules of $G_2$ in terms of granules of $G_1$. For example, <!--due to the uneven length of months, we need to use the altering-tick operation to compute months from days. We can --> consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April, June, September and November by 1 day, and shrink February by 3 days [@Ning2002-tf]. For more variations of calendar algebra operations, see @Ning2002-tf.

@@

<!-- CONTENT CUT -->
<!-- The grouping operation: -->
<!-- : Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$, -->
<!-- $$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$ -->
<!-- Example: $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$ -->

@@

<!-- CONTENT CUT -->
<!-- The altering-tick operation: -->
<!-- : Let $G_1$, $G_2$ be full-integer labeled granularities, and $l$, $k$, $m$ integers, where $G_2$ partitions $G_1$, and $1 \leq l \leq m$. The altering-tick operation $\text{Alter}^{m}_{l,k} (G_2, G_1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of $G_1$ in terms of granules of $G_2$. The altering-tick operation modifies the granules of $G_1$ so that the $l^{\text{th}}$ granule of each group has $k$ additional granules of $G_2$. -->

<!-- For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days. We can consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days [@Ning2002-tf]. For more variations of calendar algebra operations, see @Ning2002-tf. -->

@@

cricket_dot_field <- cricket_season %>%
  mutate(
    fielding_proxy = if_else(dismissal_kind %in%
      c("caught", "caught and bowled"), 1, 0),
    dot_ball_proxy = if_else(total_runs == 0, 1, 0),
    wicket_proxy = if_else(is.na(dismissal_kind), 0, 1)
  ) %>%
  group_by(
    season,
    match_id,
    batting_team,
    bowling_team,
    inning,
    over
  ) %>%
  summarise(
    runs_per_over = sum(total_runs),
    run_rate = sum(total_runs) / length(total_runs),
    fielding_wckts = sum(fielding_proxy),
    dot_balls = sum(dot_ball_proxy)
  ) %>%
  mutate(diff_run_rate = c(0, diff(run_rate)))

@@

# filter(fielding_wckts %in% c(0,1)) %>%
cricket_data %>%
  filter(over!=1) %>% 
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "boxplot",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)
) + ggtitle("") + scale_y_log10()
#+ geom_boxplot(width = 0.5, aes(colour = lag_field)) 
```

@@

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Toscas [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/).

>>>>

@@

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a _bottom granularity_ and calendar algebra operations are designed to generate new granularities recursively from the bottom up. Some relevant calendar algebra operations are discussed below; these will be used in Section \ref{sec:cyclic-gran} for illustrations in circular and aperiodic cyclic granularities.

<!-- CONTENT CUT -->
<!-- For a coarser granularity $G_2$ and a finer one $G_1$, a _grouping operation_ partitions some granules of $G_1$  to form a granule of $G_2$. For example, $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$. The _altering-tick operation_ generates new full-integer labeled granularity by periodically expanding or shrinking granules of $G_2$ in terms of granules of $G_1$. For example, <!--due to the uneven length of months, we need to use the altering-tick operation to compute months from days. We can consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April, June, September and November by 1 day, and shrink February by 3 days [@Ning2002-tf]. For more variations of calendar algebra operations, see @Ning2002-tf. -->

@@

The grouping operation:
: Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
Example: $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$

@@

The altering-tick operation:
: Let $G_1$, $G_2$ be full-integer labeled granularities, and $l$, $k$, $m$ integers, where $G_2$ partitions $G_1$, and $1 \leq l \leq m$. The altering-tick operation $\text{Alter}^{m}_{l,k} (G_2, G_1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of $G_1$ in terms of granules of $G_2$. The altering-tick operation modifies the granules of $G_1$ so that the $l^{\text{th}}$ granule of each group has $k$ additional granules of $G_2$.

For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days. We can consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days [@Ning2002-tf]. For more variations of calendar algebra operations, see @Ning2002-tf.

@@

cricket_dot_field <- cricket_season %>%
  mutate(
    fielding_proxy = if_else(dismissal_kind %in%
      c("caught", "caught and bowled"), 1, 0),
    dot_ball_proxy = if_else(total_runs == 0, 1, 0),
    wicket_proxy = if_else(is.na(dismissal_kind), 0, 1)
  ) %>%
  group_by(
    season,
    match_id,
    batting_team,
    bowling_team,
    inning,
    over
  ) %>%
  summarise(
    runs_per_over = sum(total_runs),
    run_rate = sum(total_runs)*6 / length(total_runs),
    fielding_wckts = sum(fielding_proxy),
    dot_balls = sum(dot_ball_proxy)
  ) %>%
  mutate(diff_run_rate = c(0, diff(run_rate)))

@@

# filter(fielding_wckts %in% c(0,1)) %>%
cricket_data %>%
  filter(over!=1) %>% 
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "boxplot",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)
) + ggtitle("") 
#+ geom_boxplot(width = 0.5, aes(colour = lag_field)) 
```

@@

# Software implementation {#sec:software}

The ideas presented in this article have been implemented in the open-source R [@R-language] package gravitas [@R-gravitas], available from CRAN.

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Toscas [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.
